---
title: "CK2"
author: "Group 10"
date: "2022-11-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(foreign)
library(mice)
data <- read.dta("fram.dta")
```

## Q4. Missing Data Analysis

### Types of Missingness
From the plot below, we see 14 different patterns of missingness. Patterns with most missing values are $\textbf{1. missing in sc1}$,  $\textbf{2.missing in sc2}$, and $\textbf{3. missing in sc1 and sc2}$. 
```{r}
md.pattern(data, rotate.names = TRUE)
```

By fitting the logistic regression of indicator `R1` with the rest of the covariate, we identify significant predictors with p-values less than 0.05. These are `cexam`, `sex`, `dbp`, `smoke`, and `sc2`. Thus we say that missingness in `sc1` is NOT MCAR. 


By fitting the logistic regression of indicator `R2` with the rest of the covariate, we identify significant predictors with p-values less than 0.05. These are `Lexam`, `dbp`, and `sbp`. Thus we say that missingness in `sc2` is NOT MCAR. 


By fitting the logistic regression of indicator `R3` with the rest of the covariate, we did NOT identify significant predictors with p-values less than 0.05. T= Thus we say that missingness in the (`sc1`, `sc2`) pair is MCAR. 


```{r}
### Create indicator variable R1 where R1=1 if sample experience missingness in sc1
data <- data %>% mutate(R1 = if_else(is.na(sc1), 1, 0))
data <- data %>% mutate(R2 = if_else(is.na(sc2), 1, 0))
data <- data %>% mutate(R3 = if_else(R1+R2==2, 1, 0))
### Test for MCAR by logistic regression
logit1 <- glm(R1 ~ lexam + surv + cause + cexam + cva + ca + oth + sex + age + dbp + sbp + ht + wt + mrw + smoke +sc2, data = data, family = "binomial")
summary(logit1)
logit2 <- glm(R2 ~ lexam + surv + cause + cexam + cva + ca + oth + sex + age + dbp + sbp + ht + wt + mrw + smoke +sc1, data = data, family = "binomial")
summary(logit2)
logit3 <- glm(R3 ~ lexam + surv + cause + cexam + cva + ca + oth + sex + age + dbp + sbp + ht + wt + mrw + smoke, data = data, family = "binomial")
summary(logit3)
```

To argue for whether missingness in `sc1` is MAR or MNAR, we reason with context. `sc1` stands for serum cholesterol exam 1, which is the serum cholesterol level of each individual from their first exam.  For `sc1` to be MNAR, it's missingness has to be dependent on `sc1` itself. This means those with higher or lower `sc1` may be less or more likely to take the examination. Since we see no logical explanation behind the former statement, we are more inclined to conclude that are data is MAR. Using the same logic, we say missingness in`sc2` is also MAR.

### Accomodation

We first examine the total number of missing values per column. The 5% threshold is 260. We see only missing values in `sc1` and `sc2` with count greater than the threshold. Thus, for columns with missing value count less than the threshold, one method would be to use the Complete Case Analysis.

```{r}
colSums(is.na(data))
sum(is.na(data$sc1)) / nrow(data)
sum(is.na(data$sc2)) / nrow(data)
nrow(data) * 0.05
```

From the pattern plot above, among the 14 patterns of missing data, only the top 4 has significant count of missing values. Hence, we usde Multivariate Imputation with 4 imputations.

To ensure our imputation did not add new information to the data, we drew correlation plots for data before (only include complete cases) and after imputation. We can see from below that the plot looks very similar.

```{r}
library(corrplot)
data <- data %>% mutate(diag = if_else(cexam == 0, 0, 1))
tempData = mice(data, m =4, "pmm", printFlag = FALSE, seed = 2022)
data_imputed <- complete(tempData, action=1)
M = cor(data_imputed[, -c(4, 19, 20, 21)])
corrplot(M, addCoef.col = 'black',  number.cex= 7/(ncol(data) - 4))
N = cor(data[complete.cases(data), -c(4, 19, 20, 21)])
corrplot(N, addCoef.col = 'black',  number.cex= 7/(ncol(data) - 4))
```
## Q5 Model fitting.

We first include a `BMI` column and drop `wt`, `ht`, and `mrw` since `BMI` is a linear combination of the dropped variables.

```{r}
data_imputed$bmi = 703 * data_imputed$wt / (data_imputed$ht)^2
```
### Poisson Modeling for diagnosis count per exam category

We first examine whether the mean=variance assumption hold

```{r}
p_data <- select(data_imputed, c('sex', 'age', 'sc1', 'sc2', 'dbp', 'sbp', 'smoke', 'bmi','cexam'))
p_data <- p_data %>% mutate(count = if_else(cexam!=0, 1, 0))
### calculate exposure time for offset, average exam interval is 3 years
p_data <- p_data %>% mutate(years = case_when(cexam == 0 ~ 15 * 3,
                                              cexam == 1 ~ 0,
                                              cexam != 0 & cexam != 1 ~ (cexam -1) * 3))
summary(p_data)
```

```{r}
library(sandwich)
library(lmtest)
p_mod <- glm(formula = count ~ sex + age + sc1 + sc2 + dbp + sbp + smoke + bmi, data= p_data, family=poisson())
robust <- coeftest(p_mod, vcov=sandwich)
robust
```